Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-05-20 analysis/paper/paper.Rmd
Looked more concretely at some of the data reported by Barton et al. and compared with Meso S-E Norway. Seems its generally less curated here - which makes sense given both the coastal and meso setting.

This study employs multivariate exploratory statistics to analyse the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway. This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time. However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g. @dibble2017; @rezek2020], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites. The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials. Finally, analytic and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined. Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern. In an attempt to limit the influence of confounding effects, the material chosen for analysis has a relatively constrained geographical distribution, and stems from relatively recent investigations that have employed comparable methods for excavation and classification.

@@

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [@glørstad2010]. Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @sørensen1979; @romundset2018]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of relative sea-level change allows for a relatively good chronological control of when these accumulation events occurred. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution as any stratigraphic relationship between the events is lost [@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year in the Late Mesolithic [@sørensen2014]. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared --- a point that is returned to in the discussion of the results.

@@

Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [e.g. @helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or been concerned with technological processes associated with certain sub-categories of the site inventories, such as blades or axes [e.g. @damlien2016; @berg-hansen2017; @solheim2020]. Some studies have involved in-depth analyses of handful of sites in terms of intrasite spatial organisation [@nærøy2018], and technological organisation [@eigeland2015; @mansrud2016;]. There are also some more comprehensive evaluations of a larger number of sites with associated lithic assemblages to get at potential site types and related mobility patterns [@breivik2016; @viken2018 ; @breivik2020]. However, the narratively driven analyses of these data makes the weighting of different variables unclear, and these studies draw their conclusions based on an arguably over-constraining predefined trichotomy of different site types. Finally, only a single study concerned with eight Middle-Mesolithic sites has employed a multivariate quantitative framework to help structure such a treatment [@solheim2013a]. These sites are also included here. In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods that makes it difficult, if not impossible, to consistently balance and structure the comparison of a larger number of sites with associated assemblages holding lithic artefacts numbering in the thousands.

@@

As the title of the paper suggest, the data has here been analysed from an exploratory perspective, meaning that a wide range of methods has been applied, and various combinations and transformations of variables has been explored. Only parts of this process can sensibly be reported. However, this kind of data dredging does arguably not constitute a good foundation for the pursuit of causal relationships, as this process is most certainly selectively reported upon and heavily impacted by my own biases. It can, however, result in the identification of interesting relationships that can be subjected to more rigorous testing using different data and approaches in subsequent studies. Furthermore, the employed data and R script is freely available as a research compendium following @marwick, allowing readers to explore and scrutinise the data and the final analytical choices made [cf. @Marwick2017].

The first part of the analysis conducted involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods. This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful. However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995]. Consequently, several artefact categories have here been collapased for the CA. This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading. These have all been combined into the single category "small flint tools". (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material). While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption is therefore that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While ultimately intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.

However, for the most part we lack even a most basic understanding of what any individual lithic object has been used for [@dibble2017]. This has major implications that the above-outlined analysis does not take properly into account, rendering it difficult to align any identified pattern with specific behavioural dimensions. For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life. While use-wear analysis could potentially offer a way to identify what artefacts were used for towards the end of their use-life, these kinds of analyses are extremely time-consuming and are therefore typically only conducted on a smaller number of artefacts that have already been selected for analysis based on their shape [e.g. @solheim2018]. Thus, while these analyses can potentially get at in-group variation pertaining to the end-state of a group of artefacts, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017]. As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with these inferential limitations in mind [see @clark2017, and below]. The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1979]. An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of tool replacement. Curated technological organisation, on the other hand, has been defined as related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@binford1979].

However, following not least from the ambiguous definition first put forward by @binford1979, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed, and no single definition has ever been reached [e.g. @bamforth1986; @shott1996; @surovell2009, 9--13]. The continuum between curated and expedient technology has for example been related to dimensions such as land-use and mobility strategies, as well as raw-material quality and availability [e.g. @parry1987; @kuhn1992; @andrefsky1994; @smith2015; @clark2017]. Still, that the distinction can offer a useful analytical point of departure if clearly and explicitly operationalised seems more or less agreed upon, and some dimensions of the concept are generally accepted. For example, although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool --- or extending its use-life --- by the repeated rejuvenation and modification of edges [e.g. @bamforth1986; @dibble1995; @shott2005]. Furthermore, one concrete operationalisation of the terms have been forwarded by Barton and collegues [e.g. -@barton1999; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages provide consistent results across a wide range of chronological and cultural context, and is a measure taken to reflect degree of curation. This is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the sites [@barton2014]. In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades [@clark2017].

@@

# Theory/calculation

@@

# Results

@@

# Find date to hold the central tendency date
centdate <- (sum(range(lithics$mean_date)) / 2) / 1000
```

@@

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, than cores themselves. The non-flint material on the earliest sites appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites. These findings are in large part in line with previous research [e.g. @bang-andersen2003], although possibly highlighting the importance of the non-flint materials to a larger extent than what has traditionally been the case.

The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles. While the outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites. Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot. This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g. @eigeland2015, 370, @glørstad2011].

@@

wabicor <- cor(log10(lithics2$lithic_dens), log10(lithics2$secondary_freq))
wabicorflint <- cor(log10(flint2$flint_dens), log10(flint2$secondary_freq))
```

@@

```{r pca-analysis, fig.cap = "PCA using variables related. Note that details on the transformation of the variables has been left out of the plot for clarity.", fig.height = 7, fig.width = 10}
# Rename the transformed variables again to avoid clutter in the plot. 
# Specify in figure caption that the transformed variables are used 
lithicspca <- lithicstr %>% 
  mutate(mean_date =  `Mean date cal. BCE*-1` * -1, .keep = "unused") %>% 
  rename(
   lithic_dens = "log10(lithic_dens)",
   secondary_freq = "log10(secondary_freq)",
   core_freq = "core_freq^(1/3)"
  )

@@

Figure \@ref(fig:wabi) clearly illustrates the negative correlation between the two variables (r = `r round(wabicor, 2)`) and a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites. Following @clark2017, this would be in line with a general transition from a expedient to curated lithic technology. As variable raw-material availability and workability can also impact these measures [@manninen2014]. Variation in raw-material use over time could consequently mean that these patterns could be an effect of the available raw material rather than mobility strategies. However, as is evident in when considering both plots in \@ref(fig:wabi), the pattern is evident for both the lithics considered in aggregate and for the flint data specifically. Although the distribution is slightly more spread out along the y-axis than the x-axis in the flint data, indicating that differences in volumtric density of flint is not as clearly temporally contingent, the general pattern is the same.

```{r wabi, fig.cap = "Relative frequency of secondarily worked lithics plotted against the volumetric density of artefacts (artefact count / excavated m^3^) for A) All lithics, B) Flint. The logarithm is taken to base 10 on all axes.", fig.height = 5, fig.width = 10}
w1 <- ggplot(lithics2, aes(lithic_dens, secondary_freq * 100, 
                     fill = mean_date / 1000)) + 
  geom_point(shape = 21, size = 3) +
  geom_smooth(method = "lm", col = "black", se = FALSE) +
  labs(title = "A",
       x = "Volumetric density of lithics (log10)", 
       y = "Relative frequency of secondarily worked lithics (log10)") +
  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  scale_y_log10() +
  scale_x_log10() +
  theme_classic()

@@

To visualise this temporal trend more clearly, a curation index based on the two variables was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1. The values for artefact density was then made negative to reflect its relationship with degree of curation. The mean was then found for each site on these two normalised values. To account for the temporal uncertainty associated with the dating of the sites, a simulation-based approach was also adopted [e.g. @crema2012; @baxter2016; @orton2017]. A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites \@ref(fig:dist). For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range. This simulation was repeated 1000 times, the results of which is visualised in \@ref(fig:simage). Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE. This stabilises around 7000 BCE without any further major fluctuations for the rest of the Mesolithic. It is also interesting to note that the variation in degree of curation is higher after 8000 BCE, potentially reflecting variation in site activities and associated mobility patterns.

@@

```{r simage, cache = TRUE, fig.cap = "Temporal variation in the curation index for A) all lithics in aggregate, and B) only the flint data. The temporal uncertainty is handled by means of a simulation approach, where the site ages are drawn from their repspective age determination probability distributions given in Figure 1B. A LOESS curve has been fit to the distribution for each of the 1000 simulation runs. Each simulation run is plotted with some transparency. Sites mentioned in the text are coloured.", fig.height = 10, fig.width = 12}

@@

  results[,i] <- agesamples
  row.names(results) <- row.names(agesamples)
}

@@

i2 <- ggplot(resultslong, aes(x = age, y = index_flint, colour = site_col)) + 
  geom_smooth(aes(group = sim), method = "loess", fill = "grey", col = "black",
              alpha = 0.05, se = TRUE) +
  geom_point(size = 0.5, alpha = 0.01) +
  scale_color_manual(name = NULL, breaks = c("Langangen Vestgård 1", 
                                             "Hegna vest 4","Gunnarsrød 6b",
                                             "Prestemoen 1",  "Hovland 3",
                                             "Vallermyrene 4"),
                    values = c("Langangen Vestgård 1" = "red",
                               "Hegna vest 4" = "blue",
                               "Gunnarsrød 6b" = "violet",
                               "Prestemoen 1" = "darkorange",
                               "Hovland 3" = "green",
                               "Vallermyrene 4"  = "cyan1", 
                               "other" = "black"),
                     guide = guide_legend(override.aes = list(size = 3,
                                                              alpha = 1))) +
  scale_x_continuous(breaks = c(seq(-9000,-4000, 1000))) +
  labs(title = "A", x = "Cal. BCE", y = "Curation index (flint)") +
  theme_classic() +
  theme(legend.position = "bottom")

@@

>>>>

This study employs multivariate exploratory statistics to analyse the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway. This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time. However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g. @dibble2017; @rezek2020], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites. The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as procurement strategies and access to raw materials. Finally, analytic and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined. Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern. In an attempt to limit the influence of confounding effects, the material chosen for analysis has a relatively constrained geographical distribution, and stems from relatively recent investigations that have employed comparable methods for excavation and classification within larger unified projects.

@@

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [@bjerck2010; @glørstad2010]. Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @sørensen1979; @romundset2018]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of relative sea-level change allows for a relatively good chronological control of when these accumulation events occurred. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution as any stratigraphic relationship between the events is lost [@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year in the Late Mesolithic [@sørensen2014]. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared --- a point that is returned to in the discussion of the results.

@@

Studies concerned with chronological changes in the composition of lithic assemblages in Norway have typically had a focus on chronological trends associated with formal tool types [e.g. @helskog1976; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g. @damlien2016; @berg-hansen2017; @solheim2020]. Studies that have involved the entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018; @breivik2020], or invovled extremely in-depth studies of technological organisation associated with a handful of assemblages [@eigeland2015; @mansrud2016]. These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear. To my knowledege, only a single study in southern Norway, concerned with the analysis of eight Middle Mesolithic assemblages, has involved the use of a multivariate quantitative framework to structure the analysis [@solheim2013a]. In sum, therefore, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods that makes it difficult to consistently balance the comparison of a larger number of assemblages.

@@

The relatively constrained geographical distribution of the analysed sites, the limited temporal spread over which they where investigated, as well as the methodological equivalency across projects hopefully leads to an exclusion of some biases that might otherwise skew an exploratory analysis, rendering it more likely that behaviourally meaningful patterns are identified. However, the exploratory perspective means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data. While only parts of this process can sensibly be reported on here, the data and employed R programming script is freely available as a research compendium following @marwick, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017]. However, this inductive data-dredging or pattern searching approach does constitute a limited inferential framework [@clark2009], as it involves what Binford termed a post hoc accommodation of explanations to meet the observed data --- data that is both selectively and subjectively reported upon. The process can still provide the identification of empirical patterns along the employed units of analysis, which in turn can form the basis for social and behavioural hypotheses and lay the foundation for a deductive research agenda for which clear test implications can be derived [@clark2009, 29].

The first part of the analysis involved employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. The purpose of this exercise was to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods. This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful. However, the approach taken was also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995]. Consequently, several artefact categories have here been collapased for the CA. This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading. These have all been combined into the single category "small flint tools". (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material). While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While largely intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.

However, for the most part we lack even a most basic understanding of what any individual lithic object in an assemblage has been used for [@dibble2017]. For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life. While use-wear analysis could potentially offer a way to identify what artefacts were used for towards the end of their use-life, these kinds of analyses are extremely time-consuming and are therefore typically only conducted on a smaller number of artefacts that have already been selected for analysis based on their shape [e.g. @solheim2018]. Thus, while these analyses can potentially get at in-group variation pertaining to the end-state of a group of artefacts, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017].  This has major implications that the above-outlined analysis does not take properly into account, rendering it difficult to align any identified pattern with specific behavioural dimensions. As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages with these inferential limitations in mind [see @barton2011; @clark2017, and below]. The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1979]. An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of tool replacement. Curated technological organisation, on the other hand, has been defined as related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@binford1979].

However, following not least from the ambiguous definition first put forward by @binford1979, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed, and no single definition has ever been reached [e.g. @bamforth1986; @shott1996; @surovell2009, 9--13]. The continuum between curated and expedient technology has for example been related to dimensions such as land-use and mobility strategies, as well as raw-material quality and availability [e.g. @parry1987; @kuhn1992; @andrefsky1994; @smith2015; @clark2017]. Still, that the distinction can offer a useful analytical point of departure if clearly and explicitly operationalised seems more or less agreed upon, and some dimensions of the concept are generally accepted. For example, although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool --- or extending its use-life --- by the repeated rejuvenation and modification of edges [e.g. @bamforth1986; @dibble1995; @shott2005]. Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1999 and collegues [e.g. @barton2011; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages provide consistent results across a wide range of chronological and cultural context, and is a measure taken to reflect degree of curation. This is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the sites [@barton2014]. In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades [@clark2017].

@@

@@

@@

# Find date to hold the central tendency date
centdate <- (sum(range(lithics$mean_date)) / 2) / 1000
```
# Results

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, than cores themselves. The non-flint material on the earliest sites appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites. These findings are in large part in line with previous research [e.g. @bang-andersen2003], although highlight the importance of non-flint materials for the production of projectiles.

The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles. The outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites. Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot. This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g. @eigeland2015, 370, @glørstad2011].

@@

@@

wabicor <- cor(log10(lithics2$lithic_dens), log10(lithics2$secondary_freq))
# Langangen Vestgård 2 does not have any retouched lithics
wabiflint <- filter(flint2, secondary_freq > 0)
wabicorflint <- cor(log10(wabiflint$flint_dens), log10(wabiflint$secondary_freq))

# Correlation with mean site date
secfreqcor <- cor(lithics2$mean_date, lithics2$secondary_freq)
denscor <- cor(lithics2$mean_date, lithics2$lithic_dens)

secfreqcorflint <- cor(wabiflint$mean_date, wabiflint$secondary_freq)
denscorflint <- cor(wabiflint$mean_date, wabiflint$flint_dens)

@@

```{r pca-analysis, fig.cap = "PCA using variables related. Note that details on the transformation of the variables has been left out of the plot for clarity."}
# , fig.height = 7, fig.width = 10

# Rename the transformed variables again to avoid clutter in the plot. 
# Specify in figure caption that the transformed variables are used 
lithicspca <- lithicstr %>% 
  mutate(mean_date =  `Mean date cal. BCE*-1` * -1, .keep = "unused") %>% 
  rename(
   lithic_dens = "log10(lithic_dens)",
   secondary_freq = "log10(secondary_freq)",
   core_freq = "core_freq^(1/3)"
  )

@@

Figure \@ref(fig:wabi)A illustrates the negative correlation between the two variables (*r = *`r round(wabicor, 2)`) while also displaying a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites. The correlation is stronger between the mean site age and RFSL (*r = *`r round(secfreqcor, 2)`), than between mean site age and VDL (*r = *`r round(denscor, 2)`).  As variable non-flint availability and workability has also been suggested to potentially impact these measures [@manninen2014], Figure \@ref(fig:wabi)B displays the same relationship, but exclusively for the flint data. While the negative correlation is slightly less marked when only the flint data is considered (*r = *`r round(wabicorflint, 2)`), the general pattern is the same. The relationship between mean site age and RFSF is even stronger (*r = *`r round(secfreqcorflint, 2)`), but as indicated by the more spread out distribution along the y-axis, the VDF is not temporally contingent (*r = *`r round(denscorflint, 2)`). This could perhaps indicate that when non-flint materials make up a higher share of the assemblages, as is especially the case for some of the later Mesolithic sites, expediency might to a larger extent be associated with the non-flint component of the assemblages rather than the flint material. This point is also returned to below where the temporal dimension of the relationship is explored further.

```{r wabi, fig.cap = "Relative frequency of secondarily worked lithics plotted against the volumetric density of artefacts (artefact count / excavated m^3^) for A) All lithics, B) Flint. The logarithm is taken to base 10 on all axes.", fig.height = 5, fig.width = 10}

w1 <- ggplot(lithics2, aes(lithic_dens, secondary_freq * 100, 
                     fill = mean_date / 1000)) + 
  geom_point(shape = 21, size = 3) +
  geom_smooth(method = "lm", col = "black", se = FALSE) +
  labs(title = "A",
       x = "Volumetric density of lithics (log10)", 
       y = "Relative frequency of secondarily worked lithics (log10)") +
  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
                        high="red", midpoint = centdate) +
  scale_y_log10() +
  scale_x_log10() +
  theme_classic()

@@

To get at this temporal trend a curation index based on the two variables was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1. The values for artefact density was then made negative to reflect its relationship with degree of curation. The mean was then found for each site on these two normalised values. To account for the temporal uncertainty associated with the dating of the sites, a simulation-based approach was also adopted [e.g. @crema2012; @baxter2016; @orton2017]. A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites as provided in Figure \@ref(fig:dist). For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range. This simulation was repeated 1000 times, the results of which is visualised in Figure \@ref(fig:simplot). Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE. This stabilises around 7000 BCE and remains stable without any major fluctuations for the rest of the Mesolithic. The variation in degree of curation is also markedly higher after 8000 BCE, potentially reflecting variation in site activities and associated mobility patterns.

@@

```{r simage, cache = TRUE}

@@

  results[,i] <- agesamples
  row.names(results) <- row.names(agesamples)
}
```

```{r simplot, cache = TRUE, fig.cap = "Temporal variation in the curation index for A) all lithics, and B) flint. The temporal uncertainty is handled by means of a simulation approach, where the site ages are drawn from their repspective age determination probability density functions given in Figure 1B. A LOESS curve has been fit to the distribution for each of the 1000 simulation runs. Each simulation run is plotted with some transparency. Sites mentioned in the text are given colour.", fig.height = 14, fig.width = 12}

@@

i2 <- ggplot(resultslong, aes(x = age, y = index_flint, colour = site_col)) + 
  geom_smooth(aes(group = sim), method = "loess", fill = "grey", col = "black",
              alpha = 0.05, se = TRUE) +
  geom_point(size = 0.5, alpha = 0.01) +
  scale_color_manual(name = NULL, breaks = c("Langangen Vestgård 1", 
                                             "Hegna vest 4","Gunnarsrød 6b",
                                             "Prestemoen 1",  "Hovland 3",
                                             "Vallermyrene 4"),
                    values = c("Langangen Vestgård 1" = "red",
                               "Hegna vest 4" = "blue",
                               "Gunnarsrød 6b" = "violet",
                               "Prestemoen 1" = "darkorange",
                               "Hovland 3" = "green",
                               "Vallermyrene 4"  = "cyan1", 
                               "other" = "black"),
                     guide = guide_legend(override.aes = list(size = 3,
                                                              alpha = 1))) +
  scale_x_continuous(breaks = c(seq(-9000,-4000, 1000))) +
  labs(title = "B", x = "Cal. BCE", y = "Curation index (flint)") +
  theme_classic() +
  theme(legend.position = "bottom")

@@

In my view this alignes well with the framework of @clark2017, suggesting that shifts in mobility patterns has caused this empirical pattern. However, the role of non-flint material, variability of flint quality. Furthermore, precisely how this mobility pattern has. Here Binfords model of briefly invoked, but there exists of course a wide range of possible configurations and mobility strategies that have subsistence data, site locations
